Liveness-Based Garbage Collection

### 2014-12-17 paper/sections/syntax-and-semantics.tex
Done -- the first version of syntax and semantics

% \paragraph{Inlining.}
{\color{blue}In spite of the ANF  restrictions it is still possible to
  inline non-recursive functions (a fact we use to prove the safety of
  liveness  analysis).  A user-function  call $(\LET\,\,  x \leftarrow
  (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$  to a function defined (after
  renaming  its  formals  and  locals  to be  disjoint  from  existing
  variables) by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by
  a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
  followed by the body $e_f$ but with its $(\RETURN\,\,w)$ expressions
  replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}

@@

The  semantics of  expressions  (and applications)  are  given by  the
judgement form $\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap',
e'$.  Here \stk\ is a stack  a list of continuation frames viewed as a
list.   Each continuation  is of  the form  $(\rho, x,  e)$  where the
location of $x$ is updated when the expression that is currently being
evaluated reaches WHNF, and $e$ is the next expression to be evaluated
in  the  environment  $\rho$.    The  start  state  is  $(\{\},[(\{\},
  \mainpgm,  \HALT)], \{\},  e_\mainpgm)$, and  the  program terminates
successfully on  reaching the  halt state $(\rho,[\,],  \heap, \HALT)$
with $\rho(\mainpgm)$ containing the  final result. Note that since we
assume that our language is not statically typed, a program with other
than syntactic errors (including those that could have been eliminated
using a static type system) lead to {\em stuck states} i.e. states not
covered by any rules in the small-step semantics.

Notation: we write $\rho[x \mapsto v]$ for the environment which is as
$\rho$ but  has value $v$ at  $x$.  We also  write $[\myvec{x} \mapsto
  \myvec{v}]$  which respectively  has  values $v_1,  \ldots, v_n$  at
$x_1,  \ldots, x_n$  and write  $[\myvec{x}  \mapsto \rho(\myvec{y})]$
when $v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.

\begin{figure}[t!]
\begin{tabular}{|c|c|c|}
\hline
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
\hline
          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
\hline
\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
[\rho(\myvec{x})/\myvec{y}], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
\hline
$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$ \\
$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
  \ell)\oplus\lfloor \rho\rfloor_{FV(S)-\{x\}})], e$} & \sc{let} \\ 
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\ 
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\ 
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
\hline
\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
  \rightsquigarrow$ \\
$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language}\label{fig:lang-semantics}
\end{figure}
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

>>>>

% \paragraph{Inlining.}
{\color{Myblue}In spite of the ANF  restrictions it is still possible to
  inline non-recursive functions (a fact we use to prove the safety of
  liveness  analysis).  A user-function  call $(\LET\,\,  x \leftarrow
  (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$  to a function defined (after
  renaming  its  formals  and  locals  to be  disjoint  from  existing
  variables) by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by
  a  sequence of  $\LET$'s  of the  form $z_i\leftarrow  (\ID\,\,y_i)$
  followed by the body $e_f$ but with its $(\RETURN\,\,w)$ expressions
  replaced by $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}

@@

The semantics of expressions (and similarly applications) are given by
the  judgement form $\rho,  \stk, \heap,  e \rightarrow  \rho', \stk',
\heap', e'$.  Here \stk\ is a stack of continuation frames represented
as  a list  in the  description of  the semantics.   Each continuation
frame  is of the  form $(\rho,  x, e)$  where the  location of  $x$ is
updated when the expression  that is currently being evaluated reaches
WHNF,  and  $e$  is  the  next  expression  to  be  evaluated  in  the
environment  $\rho$.   The  start  state is  $(\{\},[(\{\},  \mainpgm,
  \HALT)], \{\}, e_\mainpgm)$, and the program terminates successfully
on  reaching   the  halt   state  $(\rho,[\,],  \heap,   \HALT)$  with
$\rho(\mainpgm)$  containing  the final  result.  Note  that since  we
assume that our language is not statically typed, a program with other
than syntactic errors (including those that could have been eliminated
using a  static type system) lead  to {\em stuck  states} i.e.  states
not covered by any rules in the small-step semantics.

The  notation $[\myvec{x} \mapsto  \myvec{\ell}]$ represents
an  environment that maps  variables $x_1,  \ldots, x_n$  to locations
$\ell_1,  \ldots,  \ell_n$.   The  notation  $\heap[\ell  \mapsto  d]$
indicates the  updation of  a heap \heap\  at $\ell$ with  $d$.  $\rho
\oplus \rho'$  represents the  environment $\rho$ shadowed  by $\rho'$
and $\lfloor \rho \rfloor_X$  represents the environment restricted to
the locations in $X$. Finally $FV(s)$ represents the free variables in
the application $s$.

As an example, consider the rules related to the application $(+~x~y)$
involving a strict binary operator.  When both $x$ and $y$ are already
in WHNF  ({\sc prim-whnf}), the  heap is updated  with the sum  of the
arguments  and  the  continuation  gives  the next  expression  to  be
evaluated.   If  either  of  the   arguments  is  not  in  WHNF  ({\sc
  prim-1-clo} and {\sc prim-2-clo}), it  is sent for evaluation, and a
continuation marks that the evaluation  of $(+~x~y)$ has to be resumed
after the argument is evaluated.

\begin{figure}[t!]
\begin{tabular}{|c|c|c|}

\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) \mapsto \kappa], e$    &  \sc{const} \\
\hline
          & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)  \rightsquigarrow
$ \\  $\rho', S, H[z \mapsto (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto d_1], e$      &  \sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (e, \rho')$ & $\rho, S, H, (\CAR~x)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (\CAR~x))\!:\!S, H, e$      &  \sc{car-clo} \\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \shortstack[c]{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$ \\
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &  \sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } (s, \rho')$ & $\rho, S, H, (+~x~y)  \rightsquigarrow
\rho'\oplus\rho, (\rho, x, (+~x~y))\!:\!S, H, e$      &  \sc{prim-2-clo} \\
\hline
\shortstack[c]{$\mathit{f}~\mbox{defined as}$ \\ $~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H, (f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &  \sc{funcall} \\
\hline
$\ell$ is a new location& \shortstack[c]{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$ \\
$[x \mapsto \ell]\oplus\rho, S, H[\ell \mapsto (s, (x \mapsto
  \ell)\oplus\lfloor \rho\rfloor_{FV(S)\setminus\{x\}})], e$} & \sc{let} \\ 
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\ 
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\ 
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow$ \\ 
$\rho'\oplus\rho, (\rho, x, (\IF~x~e_1~e_2))\!:\!S, H,  e$} & \sc{if-clo} \\ 
\hline
\shortstack[c]{$H(\rho(x))~\mbox{is}$ \\$\mbox{whnf with value}~v$}& $\rho, (\rho', z, e)\!:\!S, H,
(\RETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ & \sc{return-whnf}\\   
\hline
$H(\rho(x)) = (e,\rho')$ & \shortstack[c]{$\rho, S, H, (\RETURN~x)
  \rightsquigarrow$ \\
$\rho'\oplus\rho,~ (\rho, x, (\RETURN~x))\!:\!S, H,  e$} & \sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language}\label{fig:lang-semantics}
\end{figure}
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
